---
title: "Data normalization"
author: "Ashwini Kumar Sharma, PhD"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print=100000)
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```

## Initialization

We start the analysis by initializing the packages required for all the analysis performed in this section. We also define the root directory, within which all the input/output operations for this project will be performed. At the [end of this document](#SI), a detailed software version information is provided for easier reproducibility of the analysis.

```{r}
library(DT)
library(tidyverse)
library(data.table)
library(WriteXLS)
library(ggrepel)
library(ggpubr)
library(patchwork)
library(pheatmap)
library(RColorBrewer)
path = "/Users/ashwin/Documents/Projects/YeastScreen/Essential_DAmP_screen/"
```

***

***

## Normalization methodology

We use two different normalization strategy to answer specific questions -

* For a given plate $i$, the Cytoplasmic and Mitochondrial roGFP2 ratios are normalized by the median of all controls in the same plate i.e for cytoplasm - $$NormCyto(i) = \frac{Cyto_i} {median(Ctrl_i)}$$  and for mitochondria, $$NormMito(i) = \frac{Mito_i} {median(Ctrl_i)}$$ 

> This type of control based plate normalization will be suitable in comparing the redox levels in across organells and nutrients since the roGFP2 ratios has been commonly normalized to the plate control.

* For a given plate $i$ with the Cytoplasmic and Mitochondrial roGFP2 ratios $j$ are normalized by their respective plate specific median roGFP2 ratios i.e for cytoplasm - $$NormCyto_{i} = \frac{Cyto_{ij} - median(Cyto_{i})} {mad(Cyto_{i})}$$ and for mitochondria, $$NormMito_{i} = \frac{Mito_{ij} - median(Mito_{i})} {mad(Mito_{i})}$$

> This type of organelle specific normalization will be suitable in comparing the redox levels in each organelle across all plates. Since this type of normalization tells us how far are the values from the plate median.

Furthermore, for each normalization strategy, we also summarized the quadruplicated values per gene by taking the median of the scaled values. Due to our outlier removal strategy in the previous section, many `NA` values were introduced. Each mutant has 4 values, we removed mutants with 2 or more `NA` values and those with only one `NA` value was substituted with the median of the other 3 observed values. Next, there were also **multiple copies of the same mutants (genes) in either the same plate or different plates**. We summarized them by taking the set of quadruplicated or median summarized values which have the maximum absolute median value.

Eventually, we generate the following -

* A table with all the replicate values per organelle and nutrient conditions and the the median summarized value from all plates
* A matrix with gene/mutants in rows and all combinations of organelle, nutrient and replicates (and median summarized value) in the column

Below is the table of cleaned raw data that we will use in our analysis.

```{r}
rawDatCleaned = readRDS(paste0(path, "data/workspaces/YeastMutantRedox_RawDataCleaned.RDS"))
rawDatCleaned_ctrlfilt = readRDS(paste0(path, "data/workspaces/YeastMutantRedox_RawData_Outliers_and_PoorControlFiltered.RDS"))

normalizeScreenData = function(normMethod)
{
  
  if(normMethod == "fracControl"){
    normDat = split(rawDatCleaned_ctrlfilt, as.character(rawDatCleaned_ctrlfilt$Type))
    normDat = lapply(normDat, function(x) {
    split(x, x$Plate)})
  }
  
  if(normMethod == "robustZ"){
    normDat = split(rawDatCleaned, rawDatCleaned$Type)
    normDat = lapply(normDat, function(x) {
    split(x, x$Plate)})
  }
  
  normDatReps = normDat
  
  for (i in 1:length(normDat))
  {
    for (j in 1:length(normDat[[i]]))
    {
      tmp = droplevels(normDat[[i]][[j]])
      
      if(nrow(tmp) > 0) #since there are no entries from plate 150, was filtered in previous step
      {
        
      #---------------------------------------------------------------------------------------------
      # Separate Controls, Cytoplasam and Mitochondrial roGFP2 ratios per plate
      #---------------------------------------------------------------------------------------------
      
      # Control
      tmp.ctrl = droplevels(tmp[tmp$Content == "Control",])
      
      # Cytoplasm
      tmp.cyto = droplevels(tmp[tmp$Content == "Cytoplasm",])
      
      # Mitochondria
      tmp.mito = droplevels(tmp[tmp$Content == "Mitochondria",])
      
      rm(tmp)
      
      #---------------------------------------------------------------------------------------------
      # Compute the corresponding normalization factors (median & median absolute deviation)
      #---------------------------------------------------------------------------------------------
      
      # Normalizing factor - Control
      nf.ctrl.med = median(tmp.ctrl$roGFP2.ratio, na.rm = T)
      
      # Normalizing factor - Cytoplasm
      nf.cyto.med = median(tmp.cyto$roGFP2.ratio, na.rm = T)
      nf.cyto.mad = mad(tmp.cyto$roGFP2.ratio, na.rm = T)
      
      # Normalizing factor - Mitochondria
      nf.mito.med = median(tmp.mito$roGFP2.ratio, na.rm = T)
      nf.mito.mad = mad(tmp.mito$roGFP2.ratio, na.rm = T)
      
      #---------------------------------------------------------------------------------------------
      # The two normalization strategies (plate control and median based)
      #---------------------------------------------------------------------------------------------
      
      # Normalization 1 - Plate control based
      if (normMethod == "fracControl")
      {
        tmp.cyto$roGFP2.ratio = tmp.cyto$roGFP2.ratio / nf.ctrl.med
        tmp.mito$roGFP2.ratio = tmp.mito$roGFP2.ratio / nf.ctrl.med
      }
      
      # Normalization 2 - Plate median based
      if (normMethod == "robustZ")
      {
        tmp.cyto$roGFP2.ratio = (tmp.cyto$roGFP2.ratio - nf.cyto.med) / nf.cyto.mad
        tmp.mito$roGFP2.ratio = (tmp.mito$roGFP2.ratio - nf.mito.med) / nf.mito.mad
      }
      
      rm(nf.ctrl.med,
         nf.cyto.med,
         nf.cyto.mad,
         nf.mito.med,
         nf.mito.mad)
      
      #---------------------------------------------------------------------------------------------
      # Normalized data with replicates - coverting from long to wide table format
      #---------------------------------------------------------------------------------------------
      
      #---Cytoplasm---#
      
      tmp.cyto.repl = tmp.cyto %>%
        select(Gene.Symbol, Plate, Group, roGFP2.ratio, Type, Content) %>%
        group_by(Gene.Symbol, Group) %>%
        mutate(pseurep = paste0("roGFP2_ratio_", 1:n())) %>%
        spread(key = pseurep, value = roGFP2.ratio) %>%
        ungroup() %>%
        select(-Group) %>%
        rename(Genes = Gene.Symbol,
               Nutrient = Type,
               Organelle = Content)
      
      #---Mitochondria---#
      
      tmp.mito.repl = tmp.mito %>%
        select(Gene.Symbol, Plate, Group, roGFP2.ratio, Type, Content) %>%
        group_by(Gene.Symbol, Group) %>%
        mutate(pseurep = paste0("roGFP2_ratio_", 1:n())) %>%
        spread(key = pseurep, value = roGFP2.ratio) %>%
        ungroup() %>%
        select(-Group) %>%
        rename(Genes = Gene.Symbol,
               Nutrient = Type,
               Organelle = Content)
      
      #--Compilation---#
      
      normDatReps[[i]][[j]] = rbind(tmp.cyto.repl, tmp.mito.repl)
      
      #Deleting
      rm(tmp.ctrl,
         tmp.cyto,
         tmp.mito,
         tmp.cyto.repl,
         tmp.mito.repl)
      }
    }
   rm(j) 
  }
  rm(i)
 
  res = lapply(normDatReps, function(x) {
    x = do.call("rbind", x)
    x = as.data.frame(x)
    x$Plate     = factor(x$Plate)
    x$Organelle = factor(x$Organelle, levels = c("Mitochondria", "Cytoplasm"))
    x$Nutrient  = factor(x$Nutrient, levels = c("Glucose", "Galactose", "Glycerol"))
    rownames(x) = 1:nrow(x)
    return(x)
  })
  
  res = do.call("rbind", res)
  
  #-------------------------------------------------------------------------------------------------
  # Summarizing mutiple mutants (i.e same gene) from the same or different plates
  # Also dropping genes with 2 or more NA values
  # For genes with just 1 NA value, replacing it with the median of the remaining 3 observed values
  #-------------------------------------------------------------------------------------------------
  
  res = res %>%
    rowwise() %>%
    mutate(Median_roGFP2_ratio = median(c(roGFP2_ratio_1, roGFP2_ratio_2, roGFP2_ratio_3, roGFP2_ratio_4), na.rm=T),
           NA_per_row = sum(is.na(c(roGFP2_ratio_1, roGFP2_ratio_2, roGFP2_ratio_3, roGFP2_ratio_4)))) %>%
    filter(NA_per_row < 2) %>%
    ungroup() %>%
    mutate_at(vars(starts_with("roGFO2_ratio_")),
     function(x) ifelse(is.na(x), .$Median_roGFP2_ratio, x)) %>% 
    select(-NA_per_row) %>% 
    group_by(Organelle, Nutrient, Genes) %>% 
    top_n(n=1, abs(Median_roGFP2_ratio)) %>%
    ungroup()
  
  #-------------------------------------------------------------------------------------------------
  # Compiling all replicates across organelles and nutrient conditions into a single matrix
  #-------------------------------------------------------------------------------------------------
  
  redoxMat = data.table(res[,c("Genes", "Nutrient", "Organelle", "roGFP2_ratio_1", "roGFP2_ratio_2", "roGFP2_ratio_3", "roGFP2_ratio_4")])
  redoxMat = dcast(redoxMat, Genes ~ Organelle + Nutrient, fun.aggregate = function(x){x}, fill=NA,
              value.var = c("roGFP2_ratio_1", "roGFP2_ratio_2", "roGFP2_ratio_3", "roGFP2_ratio_4"))
  redoxMat = redoxMat[,c(1,
               2,8,14,20,
               3,9,15,21,
               4,10,16,22,
               5,11,17,23,
               6,12,18,24,
               7,13,19,25
               )]
  redoxMat = data.frame(redoxMat, stringsAsFactors = F )
  rownames(redoxMat) = redoxMat$Genes
  redoxMat = redoxMat[,-1]
  
  #-------------------------------------------------------------------------------------------------
  # Compiling the median values across organelles and nutrient conditions into a single matrix
  #-------------------------------------------------------------------------------------------------
  
  redoxMat_median = data.table(res[,c("Genes", "Nutrient", "Organelle", "Median_roGFP2_ratio")])
  redoxMat_median = dcast(redoxMat_median, Genes ~ Organelle + Nutrient, fun.aggregate = function(x){x}, fill=NA,
                    value.var = "Median_roGFP2_ratio")
  
  redoxMat_median = data.frame(redoxMat_median, stringsAsFactors = F )
  rownames(redoxMat_median) = redoxMat_median$Genes
  redoxMat_median = redoxMat_median[,-1]
  
  #-------------------------------------------------------------------------------------------------
  # Putting all the results into one
  #-------------------------------------------------------------------------------------------------
  
  res = list(redox_table = res, redox_replicates = redoxMat, redox_median = redoxMat_median)
  rm(normDat, normDatReps, redoxMat, redoxMat_median)
  return(res)
}

normDat.ctrNorm = normalizeScreenData(normMethod = "fracControl")
normDat.Znorm   = normalizeScreenData(normMethod = "robustZ")
```

***

## Normalized data overview

We normalize the data as described above, below is the summary and the distribution of the normalized data -

* For per plate normalization based on plate specific control

```{r}
summary(normDat.ctrNorm$redox_table)
```

* For per plate normalization based on median organelle specific values

```{r}
summary(normDat.Znorm$redox_table)
```

* Number of mutants per condition

```{r}
a = normDat.ctrNorm$redox_table
a = split(a$Genes, paste0(a$Nutrient, a$Organelle))
sapply(a, function(x) length(unique(x)))
```

```{r, echo=F}
rm(a)
```

* Similarity between the normalization strategies

```{r}
plotList = vector("list", 6)
names(plotList) = c("Glucose-Cytoplasm",   "Glucose-Mitochondria",
                    "Galactose-Cytoplasm", "Galactose-Mitochondria",
                    "Glycerol-Cytoplasm",  "Glycerol-Mitochondria")

plotListTop = plotList

for(i in c("Glucose", "Galactose", "Glycerol"))
{
  for(j in c("Cytoplasm", "Mitochondria"))
  {
    a = normDat.Znorm$redox_table[which(normDat.Znorm$redox_table$Nutrient == i & normDat.Znorm$redox_table$Organelle == j),]
    a = a[, c("Genes", "Median_roGFP2_ratio")]
    colnames(a) = c("Genes", "Znorm")
    
    b = normDat.ctrNorm$redox_table[which(normDat.ctrNorm$redox_table$Nutrient == i & normDat.ctrNorm$redox_table$Organelle == j),]
    b = b[, c("Genes", "Median_roGFP2_ratio")]
    colnames(b) = c("Genes", "Ctrlnorm")
    
    df = merge(a,b)
    
    topZ = quantile(df$Znorm, probs = c(0.05, 0.95))
    topC = quantile(df$Ctrlnorm, probs = c(0.05, 0.95))
    
    df_top = df[which(df$Znorm < topZ[1] | df$Znorm > topZ[2] | df$Ctrlnorm < topC[1] | df$Ctrlnorm > topC[2]),]
    
    id = paste(i,j,sep="-")
    
    plotList[[id]] = ggscatter(df, x = "Znorm", y = "Ctrlnorm",
                               color = "black", shape = 20, size = 0.5, # Points color, shape and size
                               add = "reg.line",  # Add regressin line
                               add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
                               conf.int = TRUE, # Add confidence interval
                               cor.coef = TRUE, # Add correlation coefficient. see ?stat_cor
                               cor.coeff.args = list(method = "spearman", label.x = -10, label.y = 1.5, label.sep = "\n"),
                               ggtheme = theme_classic(base_size = 8)) + labs(subtitle = id)
    
     plotListTop[[id]] = ggscatter(df_top, x = "Znorm", y = "Ctrlnorm",
                               color = "black", shape = 20, size = 0.5, # Points color, shape and size
                               add = "reg.line",  # Add regressin line
                               add.params = list(color = "blue", fill = "lightgray"), # Customize reg. line
                               conf.int = TRUE, # Add confidence interval
                               cor.coef = TRUE, # Add correlation coefficient. see ?stat_cor
                               cor.coeff.args = list(method = "spearman",  label.x = -10, label.y = 1.5, label.sep = "\n"),
                               ggtheme = theme_classic(base_size = 8)) + labs(subtitle = id)
    
    
    rm(a, b, df,df_top, topZ, topC, id)
  }
  rm(j)
}
rm(i)
```

Plotting the correlation between the two normalization techniques for all matching mutants

```{r}
plotList$`Glucose-Cytoplasm` + plotList$`Glucose-Mitochondria` + 
plotList$`Galactose-Cytoplasm` + plotList$`Galactose-Mitochondria` + 
plotList$`Glycerol-Cytoplasm` + plotList$`Glycerol-Mitochondria` + 
plot_layout(nrow = 3, ncol = 2)
```

Plotting the correlation between the two normalization techniques using ONLY the top hits 5% (high/low) roGFP2 ratios from either normalization method

```{r}
plotListTop$`Glucose-Cytoplasm` + plotListTop$`Glucose-Mitochondria` + 
plotListTop$`Galactose-Cytoplasm` + plotListTop$`Galactose-Mitochondria` + 
plotListTop$`Glycerol-Cytoplasm` + plotListTop$`Glycerol-Mitochondria` + 
plot_layout(nrow = 3, ncol = 2)

```


***

## Sample similarity - Dimensionality reduction

Next, we apply the dimensionality reduction method **Multidimesional scaling (MDS)** on the roGFP2 ratio values *normalized by the plate control* and *Robust Z normalized* to identify the major grouping of the yeast mutants based on their redox status across organelle and nutrient conditions. We distinctly see that the redox status is different between the organelles (also seen by density plots, higher in mitochondria compared to cytoplasm) and within the organelles the mutants group by the nutrient conditions. 

```{r,  fig.width=7, fig.height=7, fig.align="center"}

getMDSdata = function(dat)
{
  mds = cmdscale(dist(t(dat)), eig = TRUE, k = 2)$points
  colnames(mds) = c("MDS1", "MDS2")
  
  col_anno = do.call("rbind",strsplit(rownames(mds), "_"))
  col_anno = col_anno[, - c(1:3)]
  colnames(col_anno) = c("Compartment", "Nutrient")
  mds = data.frame(mds, col_anno)
}

mdsZ = getMDSdata(dat = normDat.Znorm$redox_replicates) 
mdsC = getMDSdata(dat = normDat.ctrNorm$redox_replicates) 

p1 = ggplot(mdsZ, aes(x=MDS1, y=MDS2)) + theme_classic(base_size = 8) + labs(subtitle = "Robust Z normalized") +
geom_point(aes(color = Nutrient, shape = Compartment), size = 2) +
scale_color_manual(values = c("#fc8d62", "#66c2a5", "#8da0cb"))

p2 = ggplot(mdsC, aes(x=MDS1, y=MDS2)) + theme_classic(base_size = 8) + labs(subtitle = "Plate control normalized") +
geom_point(aes(color = Nutrient, shape = Compartment), size = 2) +
scale_color_manual(values = c("#fc8d62", "#66c2a5", "#8da0cb"))

pA = p1 + p2 + plot_layout(guides = "collect")

p3 <- ggviolin(normDat.Znorm$redox_table, x = "Organelle", y = "Median_roGFP2_ratio", xlab = "", ylab = "Median roGFP2 ratio", draw_quantiles = 0.5, fill = "Organelle", facet.by = "Nutrient", color = "grey90") + stat_compare_means(label = "p.format", method = "wilcox", cex = 2) + labs(subtitle = "Robust Z normalized") + theme_classic(base_size = 8) + theme(legend.position = "right", axis.text.x = element_blank(), axis.ticks.x = element_blank()) + scale_fill_manual(values = c("#b2df8a", "#7570b3"))


p4 <- ggviolin(normDat.ctrNorm$redox_table, x = "Organelle", y = "Median_roGFP2_ratio", xlab = "", ylab = "Median roGFP2 ratio", draw_quantiles = 0.5, fill = "Organelle", facet.by = "Nutrient", color = "grey90") + stat_compare_means(label = "p.format", method = "wilcox", cex = 2) + labs(subtitle = "Plate control normalized") + theme_classic(base_size = 8) + theme(legend.position = "right", axis.text.x = element_blank(), axis.ticks.x = element_blank()) + scale_fill_manual(values = c("#b2df8a", "#7570b3"))

pB = p3 + p4 + plot_layout(guides = "collect")

p5 = ggviolin(normDat.Znorm$redox_table, x = "Nutrient", y = "Median_roGFP2_ratio", xlab = "", ylab = "Median roGFP2 ratio", draw_quantiles = 0.5, fill = "Nutrient", facet.by = "Organelle", color = "grey90") + stat_compare_means(method = "anova", label.y = 22, label.x = 1.5, cex = 2) + stat_compare_means(label = "p.signif", method = "wilcox", ref.group = ".all.", hide.ns = TRUE) + scale_fill_manual(values = c( "#66c2a5", "#fc8d62", "#8da0cb")) + labs(subtitle = "Robust Z normalized") + theme_classic(base_size = 8) + theme(legend.position = "right", axis.text.x = element_blank(), axis.ticks.x = element_blank()) 

p6 = ggviolin(normDat.ctrNorm$redox_table, x = "Nutrient", y = "Median_roGFP2_ratio", xlab = "", ylab = "Median roGFP2 ratio", draw_quantiles = 0.5, fill = "Nutrient", facet.by = "Organelle", color = "grey90") + stat_compare_means(method = "anova", label.y = 5, label.x = 1.5, cex = 2) + stat_compare_means(label = "p.signif", method = "wilcox", ref.group = ".all.", hide.ns = TRUE) + scale_fill_manual(values = c( "#66c2a5","#fc8d62", "#8da0cb")) + labs(subtitle = "Plate control normalized") + theme_classic(base_size = 8) + theme(legend.position = "right", axis.text.x = element_blank(), axis.ticks.x = element_blank()) 

pC = p5 + p6 + plot_layout(guides = "collect")

p = pA / pB / pC

ggsave(filename = paste0(path,"analysis/normalization/roGFP2_ratio_comparison_nutrient_compartments.pdf"), plot = p, width = 7, height = 7)

p

```

```{r, echo=F}
rm(mdsZ, mdsC, p1, p2, p3, p4, p5, p6, pA, pB, pC, p)
```

***

## Replicate similarity - Correlation

Below we show the correlation among replicates.

* first with plate median (nutrient specific) based Robust Z normalization

```{r, fig.width = 5, fig.height = 5, fig.align="center"}

c1 = cor(normDat.Znorm$redox_replicates, method = "spearman", use = "pairwise.complete.obs")
c2 = cor(normDat.ctrNorm$redox_replicates, method = "spearman", use = "pairwise.complete.obs")

if(identical(colnames(c1), colnames(c2))){
  col_anno = do.call("rbind",strsplit(colnames(c1), "_"))
  col_anno = col_anno[, - c(1:3)]
  colnames(col_anno) = c("Compartment", "Nutrient")
  rownames(col_anno) = colnames(c1)
  col_anno = data.frame(col_anno, stringsAsFactors = F)
}

colr = list(Compartment = c(Mitochondria = "#b2df8a", Cytoplasm = "#7570b3"),
            Nutrient = c(Glucose = "#66c2a5", Galactose = "#fc8d62", Glycerol = "#8da0cb"))

c1[c1 == 1] = NA
c2[c2 == 1] = NA

pheatmap(c1, clustering_method = "ward.D2", 
         color = brewer.pal(n = 9, "Greys"),
         show_rownames = F, show_colnames = F,
         border_color = "white", na_col = "white",
         annotation_row = col_anno, 
         annotation_col = col_anno, 
         annotation_colors = colr, 
         fontsize = 8, main = "Robust Z normalize")
         
pheatmap(c1, clustering_method = "ward.D2", 
         color = brewer.pal(n = 9, "Greys"),
         show_rownames = F, show_colnames = F,
         border_color = "white", na_col = "white",
         annotation_row = col_anno, 
         annotation_col = col_anno, 
         annotation_colors = colr, 
         fontsize = 8, 
         filename = paste0(path,"analysis/normalization/replicate_correlation_normZnorm.pdf"), 
         width=5, height=5)
```

* second with control based normalization

```{r, fig.width = 5, fig.height = 5, fig.align="center"}

pheatmap(c2, clustering_method = "ward.D2", 
         color = brewer.pal(n = 9, "Greys"),
         show_rownames = F, show_colnames = F,
         border_color = "white", na_col = "white",
         annotation_row = col_anno, 
         annotation_col = col_anno, 
         annotation_colors = colr, 
         fontsize = 8, main = "Control normalize")

pheatmap(c2, clustering_method = "ward.D2", 
         color = brewer.pal(n = 9, "Greys"),
         show_rownames = F, show_colnames = F,
         border_color = "white", na_col = "white",
         annotation_row = col_anno, 
         annotation_col = col_anno, 
         annotation_colors = colr, 
         fontsize = 8, 
         filename = paste0(path,"analysis/normalization/replicate_correlation_normCtr.pdf"), 
         width=5, height=5)

```

```{r, echo=F}
rm(c1, c2, colr)
```

***

## Distribution of normalized data

* For per plate normalization based on plate specific control and an interactive data table to access the data

```{r, fig.height = 15, fig.width = 15}

p = ggplot(normDat.ctrNorm$redox_table) + theme_bw(base_size = 8) +
geom_boxplot(aes(x = Plate, y = Median_roGFP2_ratio), outlier.size = 0.1, lwd=0.2) + #ylim(-10, 10) +
facet_grid(Nutrient ~ Organelle) + 
theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
geom_text_repel(data = subset(normDat.ctrNorm$redox_table, Median_roGFP2_ratio > 2.5 | Median_roGFP2_ratio < 1), aes(x = Plate, y = Median_roGFP2_ratio, label = Genes), size = 2)

ggsave(filename = paste0(path,"analysis/normalization/roGFP2_ratio_distribution_PlateControl_normalized.pdf"), 
       plot = p, width = 10, height = 15)
p

tmp = normDat.ctrNorm$redox_table
tmp[,5:9] = round(tmp[,5:9],3)

datatable(tmp, rownames = FALSE, filter="top", class="compact",
          extensions = c('Buttons') , 
          options = list(autoWidth = TRUE,
                          dom = 'Bfrtip', 
                          buttons = c('csv', 'excel')
                         ))

```

```{r, echo=F}
rm(tmp, p)
```

* For per plate normalization based on median organelle specific values and an interactive data table to access the data

```{r, fig.width = 15, fig.height = 15}
p = ggplot(normDat.Znorm$redox_table) + theme_bw(base_size = 9) + #geom_hline(yintercept = c(-5,5)) +
geom_boxplot(aes(x = Plate, y = Median_roGFP2_ratio), outlier.size = 0.1, lwd = 0.2) + #ylim(-10, 10) +
facet_grid(Nutrient ~ Organelle) + 
theme(axis.text.x = element_text(angle = 90, hjust = 0.5, vjust = 0.5)) +
geom_text_repel(data = subset(normDat.Znorm$redox_table, Median_roGFP2_ratio > 3 | Median_roGFP2_ratio < -3), aes(x = Plate, y = Median_roGFP2_ratio, label = Genes), size = 2)

ggsave(filename = paste0(path,"analysis/normalization/roGFP2_ratio_distribution_RobustZ_normalized.pdf"), 
       plot = p, width = 10, height = 10)
p

tmp = normDat.Znorm$redox_table
tmp[,5:9] = round(tmp[,5:9],3)

datatable(tmp, rownames = FALSE, filter="top", class="compact",
          extensions = c('Buttons') , 
          options = list(autoWidth = TRUE,
                          dom = 'Bfrtip', 
                          buttons = c('csv', 'excel')
                         ))
```

```{r, echo=F}
rm(tmp, p)
```

***
## Saving the data

Finally we save the normalized data as a `.RDS` (R data object) and `.xlsx` excel data file. All our follow up downstream analysis will start from these normalized data.

```{r}
saveRDS(normDat.ctrNorm, paste0(path,"data/workspaces/YeastMutantRedox_NormalizedData_PlateControl.RDS"))
saveRDS(normDat.Znorm, paste0(path,"data/workspaces/YeastMutantRedox_NormalizedData_RobustZ.RDS"))

WriteXLS(normDat.ctrNorm$redox_table, ExcelFileName = paste0(path,"analysis/supplementary/tables/YeastMutantRedox_NormalizedData_PlateControl.xlsx"), 
AdjWidth =TRUE, BoldHeaderRow = TRUE, FreezeRow = 1)

WriteXLS(normDat.Znorm$redox_table, ExcelFileName = paste0(path,"analysis/supplementary/tables/YeastMutantRedox_NormalizedData_RobustZ.xlsx"), 
AdjWidth =TRUE, BoldHeaderRow = TRUE, FreezeRow = 1)

```

## Session information {#SI}
```{r}
sessionInfo()
```
