---
title: "Enrichment analysis - KEGG"
author: "Ashwini Kumar Sharma, PhD"
date: "`r Sys.Date()`"
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: hide
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)

## Global options
options(max.print=100000)
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=80)
```

## Initialization

We start the analysis by initializing the packages required for all the analysis performed in this section. We also define the root directory, within which all the input/output operations for this project will be performed. At the [end of this document](#SI), detailed software version information is provided for easier reproducibility of the analysis.

```{r}
library(hypeR)
library(reshape2)
library(pheatmap)
library(RColorBrewer)
library(dynamicTreeCut)
library(DT)

path = "/Users/ashwin/Documents/Projects/YeastScreen/Essential_DAmP_screen/"

normDat = readRDS(paste0(path,"data/workspaces/YeastMutantRedox_NormalizedData_RobustZ.RDS"))
normDat = normDat$redox_table
genesets = readRDS(paste0(path,"data/workspaces/YeastGeneSets.RDS"))

```

***

## Hypergeometric test

Here we write a simple hypergeometric test function to perform enrichment analysis. This function is based on the explainations provided on these three sources  [1](http://pedagogix-tagc.univ-mrs.fr/courses/ASG1/practicals/go_statistics_td/go_statistics_td_2015.html), [2](http://marray.economia.unimi.it/2007/material/day4/Lecture7.pdf) and [3](http://mengnote.blogspot.com/2012/12/calculate-correct-hypergeometric-p.html)


```{r}

computeEnrichment = function(gset,
                             genes,
                             fdr_cutoff = 1,
                             pval_cutoff = 1,
                             sort_output = TRUE)
{
  # removing possible duplicated entries
  gset  = lapply(gset, unique)
  genes = unique(genes)
  
  # constructing the background population of genes having at least one functional annotation
  univ  = unique(unlist(gset))
  
  # from the selection of genes, considering only those having at least one functional annotation
  genes = genes[genes %in% univ]
  
  # size of background population - i.e total number of genes with at least one functional annotation
  N = length(univ)
  
  # size of selection - total number of genes in our selection subset
  k = length(genes)
  
  # total number of genes in the functional group/pathway
  m = sapply(gset, length)
  
  # total number of genes NOT in the functional group/pathway
  n = N - m
  
  # total number of the selected genes in the functional group/pathway
  x = sapply(gset, function(y) sum(genes %in% y))
  
  # compiling final result
  res = data.frame(matrix(ncol = 6, nrow = length(gset)), stringsAsFactors = F)
  colnames(res) = c("pathway",
                    "pval",
                    "fdr",
                    "foldEnrich",
                    "overlapPercent",
                    "overlapGenes")
  
  res$pathway = names(gset)
  res$pval = signif(phyper(
    q = x - 1,
    m = m,
    n = n,
    k = k,
    lower.tail = FALSE
  ), 2)
  res$fdr = signif(p.adjust(res$pval, method = "fdr"), 2)
  res$foldEnrich = round(c((x / k) / (m / N)), 2)
  res$overlapPercent = round(c(x / m * 100), 2)
  res$overlapGenes = sapply(gset, function(x)
    paste(x[x %in% genes], collapse = "|"))
  
  # selection of significant hits
  res = res[res$pval <= pval_cutoff, ]
  res = res[res$fdr <= fdr_cutoff, ]
  
  # sort results
  if(sort_output){res = res[order(res$foldEnrich, decreasing = T), ]}
  
  # delete variables
  rm(N, k, m, n, x, univ, genes, gset)
  
  # return function output
  return(res)
}

```

***

## Enrichment analysis of top hits

Next we check the enrichment of the -

* mutants having the top and lowest 5% of roGFP2 ratios in cytoplasm with carbon source as glucose, galactose and glycerol
* mutants having the top and lowest 5% of roGFP2 ratios in mitochondria with carbon source as glucose, galactose and glycerol

across KEGG pathways

```{r}

#Cytoplasm
cyto = normDat[normDat$Organelle == "Cytoplasm",]
cyto = split(cyto, cyto$Nutrient)

cytoOX = lapply(cyto, function(x){
                                  genes = x$Genes
                                  vals = x$Median_roGFP2_ratio
                                  qt.top = quantile(vals, 0.95)
                                  selgenes = unique(genes[vals > qt.top])
                                  
                                  x = computeEnrichment(gset = genesets$KEGG, genes = selgenes)

                                  # Exactly same as in hypeR package
                                  # For sanity check of my own implemented function !!
                                  
                                  # x = hypeR(signature = selgenes[selgenes %in% unique(unlist(genesets$KEGG))],
                                  #           gsets = genesets$KEGG,
                                  #           test = "hypergeometric",
                                  #           bg = unique(unlist(genesets$KEGG)),
                                  #           pval_cutoff = 0.05,
                                  #           fdr_cutoff = 0.25,
                                  #           verbose = T)$data
                                  
                                  return(x)
                                  })

cytoRX = lapply(cyto, function(x){
                                  genes = x$Genes
                                  vals = x$Median_roGFP2_ratio
                                  qt.low = quantile(vals, 0.05)
                                  selgenes = unique(genes[vals < qt.low])
                                  
                                  x = computeEnrichment(gset = genesets$KEGG, genes = selgenes)
                                  
                                  return(x)
                                  })

#Mitochondria
mito = normDat[normDat$Organelle == "Mitochondria",]
mito = split(mito, mito$Nutrient)

mitoOX = lapply(mito, function(x){
                                  genes = x$Genes
                                  vals = x$Median_roGFP2_ratio
                                  qt.top = quantile(vals, 0.95)
                                  selgenes = unique(genes[vals > qt.top])
                                  
                                  x = computeEnrichment(gset = genesets$KEGG, genes = selgenes)
                                   
                                  return(x)
                                  })
                                  
mitoRX = lapply(mito, function(x){
                                  genes = x$Genes
                                  vals = x$Median_roGFP2_ratio
                                  qt.low = quantile(vals, 0.05)
                                  selgenes = unique(genes[vals < qt.low])
                                  
                                  x = computeEnrichment(gset = genesets$KEGG, genes = selgenes)
                                   
                                  return(x)
                                  })

rm(cyto, mito)
```

***

We construct a table containing the enrichment analysis results across all organelle and nutrient conditions.

```{r}

combo_enrichmat = rbind(cbind(cytoOX$Glucose, Organelle = "Cytoplasm", Nutrient = "Glucose", roGFP2 = "High"),
                        cbind(cytoOX$Galactose, Organelle = "Cytoplasm", Nutrient = "Galactose", roGFP2 = "High"),
                        cbind(cytoOX$Glycerol, Organelle = "Cytoplasm", Nutrient = "Glycerol", roGFP2 = "High"),
                        
                        cbind(mitoOX$Glucose, Organelle = "Mitochondria", Nutrient = "Glucose", roGFP2 = "High"),
                        cbind(mitoOX$Galactose, Organelle = "Mitochondria", Nutrient = "Galactose", roGFP2 = "High"),
                        cbind(mitoOX$Glycerol, Organelle = "Mitochondria", Nutrient = "Glycerol", roGFP2 = "High"),
                        
                        cbind(cytoRX$Glucose, Organelle = "Cytoplasm", Nutrient = "Glucose", roGFP2 = "Low"),
                        cbind(cytoRX$Galactose, Organelle = "Cytoplasm", Nutrient = "Galactose", roGFP2 = "Low"),
                        cbind(cytoRX$Glycerol, Organelle = "Cytoplasm", Nutrient = "Glycerol", roGFP2 = "Low"),
                        
                        cbind(mitoRX$Glucose, Organelle = "Mitochondria", Nutrient = "Glucose", roGFP2 = "Low"),
                        cbind(mitoRX$Galactose, Organelle = "Mitochondria", Nutrient = "Galactose", roGFP2 = "Low"),
                        cbind(mitoRX$Glycerol, Organelle = "Mitochondria", Nutrient = "Glycerol", roGFP2 = "Low")
                      )

combo_enrichmat = combo_enrichmat[,c(1:5,7,8,9,6)]

datatable(combo_enrichmat, rownames = FALSE, 
          filter = 'top', class='compact',
          extensions = c('Buttons') , 
          options = list( autoWidth = TRUE,
                          dom = 'Bfrtip', 
                          buttons = c('csv', 'excel', 'print')
                         )) %>% formatStyle("pathway","white-space"="nowrap")

```

## Visualization of enrichment analysis

We plot the following heatmaps highlighting -

* A binary matrix of the enrichment p values (fdr pvalue < 0.25 = 1 and fdr pvalue > 0.25 = 0) across all carbon sources and organelles is created. (clustering of rows by binary distance)
* Fold enrichment over background (clustering of rows by euclidean distance)
* Fraction overlap  (clustering of rows by euclidean distance)

All clustering is done by `ward.D2` method. We used the Hybrid Adaptive Tree Cutting of dendrograms provided in [dynamicTreeCut package](https://cran.r-project.org/web/packages/dynamicTreeCut/index.html) to identify optimal number of clusters. The associated paper describing the method can be found [here](https://www.ncbi.nlm.nih.gov/pubmed/18024473). Below is the heatmap showing the enrichment fdr p values (shown in purple hues, darker color corresponds to lower pvalue, fdr pvalue > 0.25 is shown in white).


```{r fig.width = 6.8, fig.height = 3, fig.align="center"}

# Creating a common matrix with all the enrichment pvalues
# Rows are pathways, columns are all combinations of nutrient, organelle and roGFP2 states

redoxEnrichMat = dcast(pathway ~ Nutrient + Organelle + roGFP2,
                       data = combo_enrichmat[, c(1, 3, 6:8)],
                       value.var = "fdr")
rownames(redoxEnrichMat) = redoxEnrichMat$pathway
redoxEnrichMat = redoxEnrichMat[, -1]

col_order = c(
  "Glucose_Mitochondria_High",
  "Galactose_Mitochondria_High",
  "Glycerol_Mitochondria_High",
  "Glucose_Cytoplasm_High",
  "Galactose_Cytoplasm_High",
  "Glycerol_Cytoplasm_High",
  "Glucose_Mitochondria_Low",
  "Galactose_Mitochondria_Low",
  "Glycerol_Mitochondria_Low",
  "Glucose_Cytoplasm_Low",
  "Galactose_Cytoplasm_Low",
  "Glycerol_Cytoplasm_Low"
)

redoxEnrichMat = redoxEnrichMat[, match(col_order, colnames(redoxEnrichMat))]
rm(col_order)

# Setting all FDR > 0.25 to NA and removing rows with all NA values
redoxEnrichMat[redoxEnrichMat > 0.25] = NA
na_cnt = apply(redoxEnrichMat, 1, function(x) {
  sum(is.na(x))
})
redoxEnrichMat = redoxEnrichMat[na_cnt < 12,]

# Performing a binary clustering on the rows of the enrichment matrix
# Binary matrix : FDR < 0.25 = 1 and FDR > 0.25 = 0
# We will use this clustering for the heatmap
redoxEnrichMat.binary = redoxEnrichMat
redoxEnrichMat.binary[is.na(redoxEnrichMat.binary)] = 0
redoxEnrichMat.binary[redoxEnrichMat.binary > 0] = 1
d = dist(redoxEnrichMat.binary, method = "binary")
h = hclust(d, method = "ward.D2")

# Identifying robust clusters using dynamic tree cutting
# We will label the rows according to these clusters
dct = cutreeHybrid(
  dendro = h,
  distM = as.matrix(d),
  minClusterSize = 5,
  verbose = 0
)
anno_rows = data.frame(Clusters = dct$labels)
rownames(anno_rows) = h$labels

# Changing the levels order such that the annotations colors and cluster bars are in same order
anno_rows_sort = anno_rows$Clusters[h$order]
level_order = unique(anno_rows_sort)
anno_rows$Clusters = factor(anno_rows$Clusters, levels = level_order)

# Find row gaps corresponding to clusters
#row_gaps = sapply(split(1:length(anno_rows_sort), anno_rows_sort), function(x)x[length(x)])
#row_gaps = row_gaps[match(levels(anno_rows$Clusters), names(row_gaps))]
#row_gaps = row_gaps[-length(row_gaps)]

# Setting the annotation order and naming for the columns
anno_columns = data.frame(
  Nutrient = rep(c("Glucose", "Galactose", "Glycerol"), 4),
  Organelle = c(
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3),
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3)
  ),
  roGFP2 = c(rep("High", 6), rep("Low", 6))
)
rownames(anno_columns) = colnames(redoxEnrichMat)

# Setting all the column and row annotation colors
col.carbon = brewer.pal(3, "Dark2")
names(col.carbon) = c("Glucose", "Galactose", "Glycerol")

col.organelle = brewer.pal(5, "Dark2")[4:5]
names(col.organelle) = c("Mitochondria", "Cytoplasm")

col.cluster =  brewer.pal(length(levels(anno_rows$Clusters)), "Paired")
names(col.cluster) = as.character(levels(anno_rows$Clusters))

anno_color = list(
  Nutrient = col.carbon,
  Organelle = col.organelle,
  roGFP2 = c(High = "black", Low = "grey"),
  Clusters = col.cluster
)

# Setting the color for the heatmap color key
color.strip = rev(colorRampPalette(brewer.pal(9, "Purples"))(6))
color.strip[length(color.strip)] = "#FFFFFF"

# Heatmap
pheatmap(
  redoxEnrichMat,
  cluster_cols = F,
  clustering_distance_rows = d,
  clustering_method = "ward.D2",
  #cutree_rows = 8,
  breaks = seq(0, 0.30, 0.05),
  fontsize = 8,
  border_color = "grey90",
  na_col = "white",
  show_colnames = F,
  gaps_col = c(3, 6, 9),
  #gaps_row = row_gaps,
  color = color.strip,
  annotation_col = anno_columns,
  annotation_row = anno_rows,
  annotation_colors = anno_color
)

```

***

Below is the heatmap showing the enrichment fold change over background. This is defined as -

* `N` be the total number of genes with at least one functional annotation (unique set of genes from all KEGG pathways),
* `k` be the total number of genes in our selection subset (say mutants with top 25% roGFP2 ratio)
* `m` be the total number of genes in the functional group/pathway (say TCA cycle)
* `x` be the the number of `k` in `m`, i.e. total number of genes common in our selection and the selected functional pathway

then, fold enrichment over background is given as $$ \frac{\frac{x} {k}} {\frac{m}{N}} $$

```{r fig.width = 6.8, fig.height = 12, fig.align="center"}

# Creating a common matrix with all the enrichment fold change
# Rows are pathways, columns are all combinations of nutrient, organelle and roGFP2 states

redoxEnrichMat = dcast(pathway ~ Nutrient + Organelle + roGFP2,
                       data = combo_enrichmat[, c(1, 4, 6:8)],
                       value.var = "foldEnrich")
rownames(redoxEnrichMat) = redoxEnrichMat$pathway
redoxEnrichMat = redoxEnrichMat[,-1]

col_order = c(
  "Glucose_Mitochondria_High",
  "Galactose_Mitochondria_High",
  "Glycerol_Mitochondria_High",
  "Glucose_Cytoplasm_High",
  "Galactose_Cytoplasm_High",
  "Glycerol_Cytoplasm_High",
  "Glucose_Mitochondria_Low",
  "Galactose_Mitochondria_Low",
  "Glycerol_Mitochondria_Low",
  "Glucose_Cytoplasm_Low",
  "Galactose_Cytoplasm_Low",
  "Glycerol_Cytoplasm_Low"
)

redoxEnrichMat = redoxEnrichMat[, match(col_order, colnames(redoxEnrichMat))]
rm(col_order)

# Row clustering
d = dist(redoxEnrichMat, method = "euclidean")
h = hclust(d, method = "ward.D2")

# Identifying robust clusters using dynamic tree cutting
# We will label the rows according to these clusters
dct = cutreeHybrid(
  dendro = h,
  distM = as.matrix(d),
  minClusterSize = 5,
  verbose = 0
)
anno_rows = data.frame(Clusters = dct$labels)
rownames(anno_rows) = h$labels

# Changing the levels order such that the annotations colors and cluster bars are in same order
anno_rows_sort = anno_rows$Clusters[h$order]
level_order = unique(anno_rows_sort)
anno_rows$Clusters = factor(anno_rows$Clusters, levels = level_order)

# Setting the annotation order and naming for the columns
anno_columns = data.frame(
  Nutrient = rep(c("Glucose", "Galactose", "Glycerol"), 4),
  Organelle = c(
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3),
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3)
  ),
  roGFP2 = c(rep("High", 6), rep("Low", 6))
)
rownames(anno_columns) = colnames(redoxEnrichMat)

# Setting all the column and row annotation colors
col.carbon = brewer.pal(3, "Dark2")
names(col.carbon) = c("Glucose", "Galactose", "Glycerol")

col.organelle = brewer.pal(5, "Dark2")[4:5]
names(col.organelle) = c("Mitochondria", "Cytoplasm")

col.cluster =  brewer.pal(length(levels(anno_rows$Clusters)), "Paired")
names(col.cluster) = as.character(levels(anno_rows$Clusters))

anno_color = list(
  Nutrient = col.carbon,
  Organelle = col.organelle,
  roGFP2 = c(High = "black", Low = "grey"),
  Clusters = col.cluster
)

# Setting the color for the heatmap color key
brk = pretty(0:round(max(redoxEnrichMat)))
color.strip = colorRampPalette(brewer.pal(9, "YlOrRd"))(c(length(brk) - 1))

# Heatmap
pheatmap(
  redoxEnrichMat,
  cluster_cols = F,
  clustering_distance_rows = d,
  clustering_method = "ward.D2",
  fontsize = 8,
  border_color = "white",
  na_col = "white",
  show_colnames = F,
  gaps_col = c(3, 6, 9),
  color = color.strip,
  annotation_col = anno_columns,
  annotation_row = anno_rows,
  annotation_colors = anno_color
)

```

***

Below is the heatmap showing the simple fraction overlap between the gene sets and the selected genes. This is defined as -


* `k` be the total number of genes in our selection subset (say mutants with top 25% roGFP2 ratio)
* `m` be the total number of genes in the functional group/pathway (say TCA cycle)
* `x` be the the number of `k` in `m`, i.e. total number of genes common in our selection and the selected functional pathway

then, fraction overlap is simply $$ \frac{x} {m} \times 100 $$

```{r fig.width = 6.8, fig.height = 12, fig.align="center"}

# Creating a common matrix with all the enrichment fold change
# Rows are pathways, columns are all combinations of nutrient, organelle and roGFP2 states

redoxEnrichMat = dcast(pathway ~ Nutrient + Organelle + roGFP2,
                       data = combo_enrichmat[, c(1, 5, 6:8)],
                       value.var = "overlapPercent")
rownames(redoxEnrichMat) = redoxEnrichMat$pathway
redoxEnrichMat = redoxEnrichMat[,-1]

col_order = c(
  "Glucose_Mitochondria_High",
  "Galactose_Mitochondria_High",
  "Glycerol_Mitochondria_High",
  "Glucose_Cytoplasm_High",
  "Galactose_Cytoplasm_High",
  "Glycerol_Cytoplasm_High",
  "Glucose_Mitochondria_Low",
  "Galactose_Mitochondria_Low",
  "Glycerol_Mitochondria_Low",
  "Glucose_Cytoplasm_Low",
  "Galactose_Cytoplasm_Low",
  "Glycerol_Cytoplasm_Low"
)

redoxEnrichMat = redoxEnrichMat[, match(col_order, colnames(redoxEnrichMat))]
rm(col_order)

# Row clustering
d = dist(redoxEnrichMat, method = "euclidean")
h = hclust(d, method = "ward.D2")

# Identifying robust clusters using dynamic tree cutting
# We will label the rows according to these clusters
dct = cutreeHybrid(
  dendro = h,
  distM = as.matrix(d),
  minClusterSize = 5,
  verbose = 0
)
anno_rows = data.frame(Clusters = dct$labels)
rownames(anno_rows) = h$labels

# Changing the levels order such that the annotations colors and cluster bars are in same order
anno_rows_sort = anno_rows$Clusters[h$order]
level_order = unique(anno_rows_sort)
anno_rows$Clusters = factor(anno_rows$Clusters, levels = level_order)

# Setting the annotation order and naming for the columns
anno_columns = data.frame(
  Nutrient = rep(c("Glucose", "Galactose", "Glycerol"), 4),
  Organelle = c(
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3),
    rep("Mitochondria", 3),
    rep("Cytoplasm", 3)
  ),
  roGFP2 = c(rep("High", 6), rep("Low", 6))
)
rownames(anno_columns) = colnames(redoxEnrichMat)

# Setting all the column and row annotation colors
col.carbon = brewer.pal(3, "Dark2")
names(col.carbon) = c("Glucose", "Galactose", "Glycerol")

col.organelle = brewer.pal(5, "Dark2")[4:5]
names(col.organelle) = c("Mitochondria", "Cytoplasm")

col.cluster =  brewer.pal(length(levels(anno_rows$Clusters)), "Paired")
names(col.cluster) = as.character(levels(anno_rows$Clusters))

anno_color = list(
  Nutrient = col.carbon,
  Organelle = col.organelle,
  roGFP2 = c(High = "black", Low = "grey"),
  Clusters = col.cluster
)

# Setting the color for the heatmap color key
brk = pretty(0:round(max(redoxEnrichMat)))
color.strip = colorRampPalette(brewer.pal(9, "BuGn"))(c(length(brk) - 1))

# Heatmap
pheatmap(
  redoxEnrichMat,
  cluster_cols = F,
  clustering_distance_rows = d,
  clustering_method = "ward.D2",
  breaks = brk,
  fontsize = 8,
  border_color = "white",
  na_col = "white",
  show_colnames = F,
  gaps_col = c(3, 6, 9),
  color = color.strip,
  annotation_col = anno_columns,
  annotation_row = anno_rows,
  annotation_colors = anno_color
)

```

## Session information {#SI}
```{r}
sessionInfo()
```




